---
title: "Governor's Challenge Notebook"
output: html_notebook
---

This notebook contains the R-version of the code for the Governor's challenge; it follows the same format as the STATA tutorial.  Let's assume you've followed the tutorial up to the point where you're starting to import the data in STATA.

I recommend using the software RStudio, because it has a nice user interface which is similar to other packages, and is fairly flexible with different R distributions.

Here, we first want to import the GDP data.  We can do this using the <b>readr</b> package, which we need to install (if necessary; uncomment these if you don't have these packages), along with a couple others)

```{r}
#install.packages("gtools")
#install.packages("readr")
#install.packages("tidyr")
#install.packages("dpylr")
#install.packages("zoo")
```

Now, we can use this package to load the data:

```{r}
library(readr)
data <- read_csv("C:/Users/Jonathan Graves/Downloads/GC2/36100434-eng/36100434.csv")
```

Next, we only want to keep the vector associated with the data we want

```{r}
data <- subset(data, VECTOR == "v65201210")
head(data)
```

Now, we can do the same thing for the CPI data (loading it and subsetting it):

```{r}
data2 <- read_csv("C:/Users/Jonathan Graves/Downloads/GC2/18100004-eng/18100004.csv")
data2 <- subset(data2, VECTOR == "v41690973")
```

Next, we want to append these datasets together, but first we have to keep only the columns they have in common

```{r}
data <- within(data, rm("Seasonal adjustment","Prices", "North American Industry Classification System (NAICS)"))
data2 <- within(data2, rm("Products and product groups"))
data <-  rbind(data,data2)
head(data)
```

Now, let's convert the <b>ref_data</b> variable into a better format:

```{r}
library(zoo)
data$DATE <- as.yearmon(data$REF_DATE)
head(data)
```

Next, let's delete all of the useless variables:

```{r}

data <- within(data, rm("REF_DATE", "UOM", "UOM_ID","SCALAR_ID", "SYMBOL", "COORDINATE","STATUS","TERMINATED","DECIMALS", "SCALAR_FACTOR"))
head(data)

```

Now, we need to reshape our dataset

```{r}
library(tidyr)
library(dplyr)
data <- spread(data, "VECTOR", "VALUE")
```

Finally, we will rename our variables and save the dataset:

```{r}
data <- rename(data, CPI = v41690973, GDP = v65201210)
head(data)
save(data,file="GDP_CPI_data.Rda")
```

# VAR Analysis

We'll assume that to this point you've been following along.  Now, we have our data stored in our `data` object.  You may want to try out the model using the other dataset, but for now it's good to see how things are different.

There's a very good reference on using VAR models in R, which you can find at https://www.econometrics-with-r.org/16-1-vector-autoregressions.html

The main element we need to do first is to format our series in `data` as time series objects.  We know that they're monthly, staring in Januart 1914 and ending in June 2019.  This will require an extra package:

```{r}

#install.packages("quantmod")
#install.packages("dynlm")
#install.packages("lmtest")
#install.packages("sandwich")
#install.packages("vars")
library(quantmod)
library(dynlm)
library(lmtest)
library(sandwich)
library(vars)

```


Next, we need to define and format our variables as time series objects, so we can compute VAR models from them.  We also want to create some difference objects:

```{r}

GDP <- ts(data$GDP[(data$DATE > "DEC 1999") & (data$DATE < "MAY 2019")], start=c(2000,1),end=c(2019,4),frequency=12)
# the frequency is 12 because this is monthly data and there are 12 months in a year
#Let's also start the analysis in 2000
logGDP <- log(GDP)

CPI <- ts(data$CPI[(data$DATE > "DEC 1999") & (data$DATE < "MAY 2019")], start=c(2000,1),end=c(2019,4),frequency=12)
logCPI <- log(CPI)

DlogGDP = 12*(logGDP - stats::lag(logGDP,n=-1))
DlogCPI = 12*(logCPI - stats::lag(logCPI,n=-1))
#we multiply by 12 here to get annualized rates

```


Now, we can make a two-variable VAR model.  First, we do it equation-by equation:

```{r}

VAR1 <- dynlm(DlogGDP ~ L(DlogGDP,1:2) + L(DlogCPI, 1:2), start = c(2000,3), end = c(2019,4))
names(VAR1$coefficients) <- c("Constant", "GDP T-1", "GDP T-2", "Inflation T-1", "Inflation T-2")

#Notice how we have to restrict the first 2 months to not enter the equation; this is because our data starts in Jan 2000, so there aren't two lags of data until Mar 2000

coeftest(VAR1, vcov. = sandwich)

VAR2 <- dynlm(DlogCPI ~ L(DlogGDP,1:2) + L(DlogCPI, 1:2), start = c(2000,3), end = c(2019,4))
names(VAR2$coefficients) <- c("Constant", "GDP T-1", "GDP T-2", "Inflation T-1", "Inflation T-2")

coeftest(VAR2, vcov. = sandwich)

```

Now, we can do it using a single VAR.  Notice how the results are the same:

```{r}

VAR_data <- window(ts.union(DlogGDP,DlogCPI), start=c(2000,1), end = c(2019,3))

VAR_est <- VAR(y = VAR_data, p = 2)
VAR_est

```

Now, let's forecast our VAR.  R automatically does an iterated forecast:

```{r}

forecast <- predict(VAR_est)

forecast
plot(forecast)
```